---
title: "StatsCheatSheet"
author: "Sydney Bell"
format: html
editor: visual
---

## Steps

1.  Load in data
2.  Decide what the variables are (categorical, numeric, continuous, etc + response, independent)
3.  Are there random effects?
4.  Centering and scaling (bglm_example4)

If you have continuous predictor, you want to center it and scale it for Bayesian analyses. Centering allows predictors to be independent of each other. It’s mean becomes 0.

In our bglm_example1, our lowest value isn’t 0 (y intercept) so we would center and use the average response (average x)

Scaling: take value and divide by standard deviation (all then have SD of 1) —> therefore, never have very large values —> these are computationally high cost
5.  Make categorical variables into factors and check their order
6.  Decide on distribution

### What’s our distribution and do our data fit assumption??

![](images/images (1).png)


7.  Exploratory data analysis
- Normal = normality (boxplot), linearity (scatterplot)
- Poisson = 
- Bernoulli = scatterplot

8. Model formula
9. Define priors 
10. Fit the model
11. Prior only model and plot conditional effects
12. Prior AND data model and plot conditional effects (priors should always be larger than posterior)
13. Visualize priors and posteriors
14. Check all diagnostics
15. Generate summary and tables
- summary() creates estimates on logit scale


## Normal/Gausian
-   measurements
-   parameters: mean and variance
-   assume: normality (boxplot), linearity (scatterplot), homogeneity of variance, independence, un-correlated predictors (multi-colinearity)

Examples: bglm_example1
Visualization: boxplot, scatterplot

### Priors
Will include slope, intercept, and sigma
- median is more robust than mean and not influenced heavily by outliers (except for zero-inflated data)
- mad is more robust than SD (median absolute deviation)
```{r}
#ask brms package for its recommended priors just to check
get_prior(fert_form, data = fert)

priors <- prior(normal("median", "mad"), class = "Intercept") +
  prior(normal(0,"mad"), class="b") +
  prior(student_t(3, 0, "mad"), class="sigma")
```


## Log Normal
-   Example: a normal model gives negative bird predictions so need to try something else like a log normal distribution

## Poisson
-   lambda = mean, variance combined
-   assuming mean/variance = 1 (dispersion), assume they are the same thing
-   need link function to match Poisson function with slope

### Priors
```{r}
#From bglm_example5

#for Intercept look at the median of ALG1 (since this is the first factor/reference and that is where the intercept comes from)
day |> 
  group_by(TREAT) |> 
  summarise(median(log(BARNACLE)), mad(log(BARNACLE))) #need to use log because we have the logit link function for a Poisson distribution (couldn't do this if we had zeros)

#intercept and all betas
priors <- prior(normal(3.1,0.3), class = 'Intercept') +
  prior(normal(0,0.4), class = 'b') #keep mean at 0 because we want to let the data decide, variance will be something big enough to include all of the mad() variances
```

### Summary
Will be generated on the log scale which is not helpful.
Need to exp.
```{r}
exp(3.08) #this is the mean # of new recruits in ALG1

#diff between ALG1 and ALG2
exp(0.26) #but when you exp() it becomes multiplicative not additive so its 1.3 times greater than 21.8
exp(3.08 + 0.26) #this gives you the mean of ALG2
(1.29 - 1) * 100 #recruitment is 29% higher in the ALG2 group than the ALG1 group
```


## Bernoulli
-   binary or proportions
-   Bernoulli is a special case of binomial.
      - Variance should equal 0
      - Don't assume linearity
-   Any event with a single trial and only two outcomes follows a Bernoulli distribution
-   parameter: pie
-   assume variance = 1

Examples: bglm_example2
Visualization: scatterplot with line to show distribution

### Priors
Slope mean will always be 0
Slope variance will be 1

Intercept mean will be 0
Intercept variance will be 1

```{r}
#set priors
priors <- prior(normal(0,1), class = 'Intercept') +
  prior(normal(0,1), class = 'b')
```

