---
title: "Bayesian GLMM Part9"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

#install.packages("effects")
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(bayesplot)
library(broom.mixed)
library(rstan)
library(patchwork)
library(DHARMa)
source('helperFunctions.R')
```



# Scenario

Once it is established that mass coral bleaching is occurring on the
Great Barrier Reef (GBR), a monitoring team is mobilised in both the
air and water in order to document the scale and extent of the
bleaching. To better understand the causes and consequences of
bleaching, one marine ecologist (lets call him Terry) was interested
in investigating differences in bleaching extent across different reef
habitats. To do so, aerial surveys were partitioned into four habitats
(C - crest, F - flank, L - lower, U - upper).

Bleaching is scored categorically according to the following scale.

| Bleaching Score | 2016 Equates to  |
|-----------------+------------------|
|               0 | No bleaching     |
|               1 | 0-10% bleaching  |
|               2 | 10-30% bleaching |
|               3 | 30-60% bleaching |
|               4 | > 60% bleaching  |

The GBR is very large and the extent of coral bleaching is not uniform
throughout the reef. Hence, Terry wanted to see if the habitat
patterns were similar throughout the GBR or whether they were
dependent on the overall bleaching severity.

# Read in the data

2016 data only

```{r readData, results='markdown', eval=TRUE}
hughes = read_csv('../data/hughes.csv', trim_ws=TRUE)
glimpse(hughes)

hughes$HABITAT |> unique()
```

| REEF   | HABITAT | SECTOR | SCORE |
|--------|---------|--------|-------|
| 09-357 | C       | North  | 4     |
| 09-357 | F       | North  | 4     |
| 09-357 | U       | North  | 3     |
|        |         |        |       |


# Data preparation
SECTOR (fixed effect) - reef can only be found in one of the sectors
REEF (random effect) - reef can only be found in one of the sectors and aren't replicated across
HABITAT (fixed effect) - can be multiple habitats in each reef
SECTOR:HABITAT (fixed effect) - interaction term

```{r}
#| label: data prep
#| eval: true
#| echo: true

#random effects are always set as factors

hughes <- hughes |> 
  mutate(REEF = factor(REEF),
         HABITAT = factor(HABITAT),
         SECTOR = factor(SECTOR, levels = c("North", "Central", "South")),
         oSCORE = factor(SCORE, ordered = TRUE)) #this allows it to be handled as an ordinal regression
```

# Exploratory Data Analysis

Proportional (cumulative link) odds-ratio models are useful when the latent
(unmeasured) response is recorded on a ordinal (ordered categorical) scale.
When this is the case, we can calculate the probability of a that an observed
ordinal score ($y$) is less than or equal to any given level ($i$: category) given a set of
covariates ($\boldsymbol{X}$) according to the following:

$$ Pr(y\le i|\boldsymbol{X}) =
\frac{1}{1+e^{-(\theta_i - \boldsymbol{\beta}.\boldsymbol{X})}}\\ $$

where $y$ is the observed categorical response, $\boldsymbol{X}$ is a ($n \times
p$) effects model matrix, $\boldsymbol{\beta}$ are the $p$ effects parameters
and $\theta_i$ are the $K-1$ category thresholds

```{r}
#| label: visualize data
#| eval: true
#| echo: true

#treat as a binomial for visual purposes (jitter. Cant look at assumptions)

#symbiont x-axis
#oSCORE y-axis
#facet by blocks

hughes |> 
  ggplot(aes(y = oSCORE, x = HABITAT)) +
  geom_point(position = position_jitter()) +
  facet_wrap(~SECTOR)

#now add lines to join up dots of each reef
hughes |> 
  ggplot(aes(y = oSCORE, x = HABITAT)) +
  geom_point(position = position_jitter()) +
  geom_line(aes(y = as.numeric(oSCORE), x = as.numeric(HABITAT), group = REEF),
            position = position_jitter()) +
  facet_wrap(~SECTOR)
#lines aren't parallel looking so consider random slopes
```


# Fit the model

::: {.panel-tabset}
```{r}
#| label: formula
#| eval: true
#| echo: true

hughes_form <- bf(oSCORE ~ HABITAT*SECTOR + (1|REEF),
                  family = cumulative(link = "logit", threshold = "flexible"))

```

```{r}
#| label: check priors
#| eval: true
#| echo: true

get_prior(hughes_form, data = hughes)
#5 intercepts, one for each SCORE level, series of binomials
#because its a series of binomials, we can't get prior info from the data
```

```{r}
#| label: set priors
#| eval: true
#| echo: true

priors <- prior(normal(0, 1), class = "Intercept") +
  prior(normal(0, 1), class = "b") +
  prior(student_t(3, 0, 1), class = "sd")

#first intercept = prob of being in cat 0 vs any above 0, for the north crest (first levels)
#second intercept = prob of being in cat 1 vs any above 1, for the north crest (first levels)
```

```{r}
#| label: fit model priors only
#| eval: true
#| echo: true

hughes_brm1 <- brm(hughes_form,
                 data=hughes,
                 prior=priors,
                 sample_prior="only",
                 iter=5000,
                 chains=3, cores=3, #one core per chain
                 warmup=1000,
                 thin=5,
                 backend="cmdstanr",
                 refresh=0,
                 seed = 1)
```

```{r}
#| label: conditional effects plot
#| eval: true
#| echo: true

hughes_brm1 |> 
  conditional_effects("HABITAT", 
                      conditions = make_conditions(hughes_brm1, "SECTOR"),
                      categorical = TRUE)
```

```{r}
#| label: fit model
#| eval: true
#| echo: true

hughes_brm2 <- update(hughes_brm1,
                 sample_prior="yes",
                 refresh=100,
                 seed = 1)

saveRDS(hughes_brm2, file = "D:/GitHub/SUYRs_public/models/hughes_brm2.rds")
hughes_brm2a <- readRDS(file = "D:/GitHub/SUYRs_public/models/hughes_brm2.rds")

#how to store multiple things
#save(hughes_brm2, hughes, hughes_form, priors, file = "hughes_analysis.RData")
#load(file = "hughes_analysis.RData")
```

```{r}
#| label: effects plot again
#| eval: true
#| echo: true

hughes_brm2 |> 
  conditional_effects("HABITAT",
                      conditions = make_conditions(hughes_brm2, "SECTOR"),
                      categorical = TRUE)
```
:::

# MCMC sampling diagnostics

::: {.panel-tabset}
```{r}
#| label: diagnostics
#| eval: true
#| echo: true

hughes_brm2$fit |> stan_trace(inc_warmup = TRUE)

#have over 10 trace plots so need to cut them down so we can see them
get_variables(hughes_brm2)
params <- get_variables(hughes_brm2) |> str_subset("^b_.*|^sd_.*")
params
#now do this
hughes_brm2$fit |> stan_trace(pars = params)

hughes_brm2$fit |> stan_ac(pars = params)

hughes_brm2$fit |> stan_rhat()

hughes_brm2$fit |> stan_ess()

hughes_brm2 |> pp_check(type = 'dens_overlay', ndraws=100) 

hughes.resids <- make_brms_dharma_res(hughes_brm2, integerResponse = TRUE)
testUniformity(hughes.resids)
plotResiduals(hughes.resids, quantreg = FALSE, asFactor = FALSE)
testDispersion(hughes.resids)
#nothing super obviously wrong
```

:::

# Model summary

::: {.panel-tabset}
```{r}
#| label: summary
#| eval: true
#| echo: true

hughes_brm2 |> summary()
plogis(-8.5) #Intercept[1] = probability of North C having a score of 0 vs a higher one
plogis(-0.4)
exp(-0.4) #odds of being in cat 3 vs cat 4 is 0.67 for North C

#HABITATF is a slope aka a difference
exp(-0.34) # change of 0.7 on a fractional scale, prob of bleaching is lower in habitat f than habitat c
```

:::

# Model investigation

::: {.panel-tabset}

```{r}
#| label: get avergaes for each group
#| eval: true
#| echo: true

#create new data
newdata <- hughes |> 
  distinct(HABITAT, SECTOR)
newdata

#asking to predict bleaching scores for the 12 combinations we just created
predict(hughes_brm2, newdata = newdata, re_formula = NA)

#weighted average = average amount of bleachin in North C is 3.535
sum(c(0.0004166667, 0.005000000, 0.05500000, 0.33833333, 0.6012500000) * 0:4)
#but we did this on the summary, not the full posterior (which would've included SD)

#If you just calculated the averages for the raw data, you're assuming a normal/gaussian distribution and all of the assumptions that come along with that
#thats why you need to do this ordinal model

#bayesian version of predict()
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) #re_formula is saying to ignore random effects, just focused on SECTOR/HABITAT level

#as a check, start with just one group/restrict the data
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  filter(.draw == 1, HABITAT == "C", SECTOR == "North")
#we would have to multiply .category by .epred to get average for each group for draw 1
#need to make .category fctors into a number first though

#try to make a new column for this calculation
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) #this prevents the cateogries from being changed to 1,2,3,4,5

epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) |> 
  mutate(score = .epred * bleaching)

#now need to add up the scores to get one average score per Habitat/Sector but we want one for each draw
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) |> 
  mutate(score = .epred * bleaching) |> 
  ungroup() |> 
  group_by(HABITAT, SECTOR, .draw) |> 
  summarise(score = sum(score))

#If we want all habitats, all sectors, and all draws...just remove the filter
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  #filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) |> 
  mutate(score = .epred * bleaching) |> 
  ungroup() |> 
  group_by(HABITAT, SECTOR, .draw) |> 
  summarise(score = sum(score))

#now need to summarise across draws
epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  #filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) |> 
  mutate(score = .epred * bleaching) |> 
  ungroup() |> 
  group_by(HABITAT, SECTOR, .draw) |> 
  summarise(score = sum(score)) |> 
  summarise_draws(median, HDInterval::hdi)
```

:::


# Further investigations

::: {.panel-tabset}
```{r}
#| label: comparisons
#| eval: true
#| echo: true

#remove summarise_draws because we want the full posteriors, always work on un-summarised data
cellmeans <- epred_draws(hughes_brm2, newdata = newdata, re_formula = NA) |> 
  #filter(.draw == 1, HABITAT == "C", SECTOR == "North") |> 
  mutate(bleaching = as.numeric(as.character(.category))) |> 
  mutate(score = .epred * bleaching) |> 
  ungroup() |> 
  group_by(HABITAT, SECTOR, .draw) |> 
  summarise(score = sum(score)) 

cellmeans |> 
  compare_levels(var = score, by = HABITAT, comparison = emmeans_comparison("pairwise"))

cellmeans |> 
  compare_levels(var = score, by = HABITAT, comparison = emmeans_comparison("pairwise")) |> 
  summarise_draws(median, HDInterval::hdi, Pl = ~mean(.x < 0), Pg = ~mean(.x > 0))

#absolute scale
#For example, habitat C is 0.1 units higher than habitat F in the North and there is strong evidence
```

```{r}
#| label: fractional scale table
#| eval: true
#| echo: true

cellmeans |> 
  mutate(lscore = log(score)) |> 
  compare_levels(var = lscore, by = HABITAT, comparison = emmeans_comparison("pairwise")) |> 
  mutate(score = exp(lscore)) |> 
  summarise_draws(median, HDInterval::hdi, Pl = ~mean(.x < 1), Pg = ~mean(.x > 1))

#bleaching is 3% higher in C North than F North
```
:::





