---
title: "Bayesian GLMM Part10"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(broom)     #for tidy output
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(bayesplot)
library(broom.mixed)
library(rstan)
library(patchwork)
library(DHARMa)
source('helperFunctions.R')
```


# Scenario

For over 35 years, the AIMS long term monitoring program has performed
benthic surveys of coral reefs on the Great Barrier Reef (GBR). To do
so the team uses two main survey techniques 1) Manta Tow and 2) Photo
transects. The current example focuses on data collected using the
later technique.
Within each reef, there are three sites on the north east flank and
within each site there are five permanent transects. Every year, the
team return and take photos every meter along the transects. Once back
in the laboratory, five points from every photo are scored according
to what is represented underneath the point.

The main objective of long-term monitoring is to be able to report on
status and trends. Specifically, what is the status of each major
benthic group (such as hard coral, soft coral and macroalgae) and how
are they changing (particularly in response to major disturbances).

For this example, we will focus on a single reef (Agincourt Reef
No.1).

# Read in the data

```{r readData, results='markdown', eval=TRUE}
ltmp <- read_csv('../data/ltmp.csv', trim_ws=TRUE)
glimpse(ltmp)
```

| AIMS_REEF_NAME | REPORT_YEAR | SITE_NO | TRANSECT_NO | HC   | n.points | total.points |
|----------------|-------------|---------|-------------|------|----------|--------------|
| Arlington Reef | 2006        | 1       | 1           | 10.0 | 20       | 200          |
| Arlington Reef | 2006        | 1       | 2           | 10.5 | 21       | 200          |
|                |             |         |             |      |          |              |


# Data preparation
Order:
REEF (random)
SITE (random)
TRANSECT (random)
REPORT_YEAR (fixed)

Could model HC (percentage) as a beta model but they are very messy and not as reliable because you could get 10% from 2/10 or 10/200 etc

```{r}
#| label: data prep
#| eval: true
#| echo: true

#categorical variables and filter to just one reef
ltmp_sub <- ltmp |> 
  dplyr::filter(AIMS_REEF_NAME == "Agincourt Reef No.1") |> 
  droplevels() |> #need to remove factors/levels that aren't included in Agincourt Reef No.1
  mutate(SITE_NO = factor(paste(AIMS_REEF_NAME, SITE_NO)),
         TRANSECT_NO = factor(paste(SITE_NO, TRANSECT_NO)),
         fYEAR = factor(REPORT_YEAR))
ltmp_sub
```

# Exploratory Data Analysis

```{r}
#| label: visualize data
#| eval: true
#| echo: true

ltmp_sub |> 
  ggplot(aes(y = HC, x = REPORT_YEAR)) +
  geom_line(aes(group = TRANSECT_NO, colour = SITE_NO)) +
  geom_point()
#linearity is not met!!!
```

# Fit the model

::: {.panel-tabset}

```{r}
#| label: model formula
#| eval: true
#| echo: true

#treat YEAR as categorical 
#The intercept would be the first year (1995) but this first year had less photos and wonky data
#so lets change the intercept to the most recent year instead using reverse
ltmp_sub <- ltmp_sub |> 
  mutate(fYEAR = factor(fYEAR, levels = rev(levels(fYEAR))))

#writing the formula this way has transect nested within site
ltmp_sub_form <- bf(n.points | trials(total.points) ~ fYEAR + (1|SITE_NO) + (1|TRANSECT_NO),
                    family = binomial(link = "logit"))

```

```{r}
#| label: priors
#| eval: true
#| echo: true

get_prior(ltmp_sub_form, data = ltmp_sub)

#intercept is the average HC at 2023
ltmp_sub |> group_by(fYEAR) |> 
  summarise(Median = median(qlogis(HC/100)), MAD = mad(qlogis(HC/100)))

#set priors
priors <- prior(normal(-0.8, 0.3), class = "Intercept") +
  prior(normal(0, 3), class = "b") +
  prior(student_t(3, 0, 0.5), class = "sd")
```

```{r}
#| label: run model
#| eval: true
#| echo: true

ltmp_sub_brm1 <- brm(ltmp_sub_form,
                 data=ltmp_sub,
                 prior=priors,
                 sample_prior="only",
                 iter=5000,
                 chains=3, cores=3, #one core per chain
                 warmup=1000,
                 thin=5,
                 backend="cmdstanr",
                 refresh=0,
                 seed = 1,
                 control = list(adapt_delta = 0.99))


ltmp_sub_brm1 |> 
  conditional_effects(conditions = data.frame(total.points = 200)) |> 
  plot(points = TRUE)
```
```{r}
#| label: run model with data
#| eval: true
#| echo: true

ltmp_sub_brm2 <- update(ltmp_sub_brm1,
                        sample_prior = "yes",
                        chains = 3, cores = 3,
                        control = list(adapt_delta = 0.99, max_treedepth = 20), #sometimes have to increase this when you have a more complex model
                        seed = 1,
                        refresh = 100)

ltmp_sub_brm2 |> 
  conditional_effects(conditions = data.frame(total.points = 200)) |> 
  plot(points = TRUE)
```
:::

# MCMC sampling diagnostics

::: {.panel-tabset}

```{r diagnostics, eval=TRUE, echo=TRUE}

ltmp_sub_brm2 |> 
  hypothesis("fYEAR2021=0") |> 
  plot()

ltmp_sub_brm2$fit |> stan_trace(inc_warmup = TRUE)

#have over 10 trace plots so need to cut them down so we can see them
get_variables(ltmp_sub_brm2)
params <- get_variables(ltmp_sub_brm2) |> str_subset("^b_.*|^sd_.*")
params
#now do this
ltmp_sub_brm2$fit |> stan_trace(pars = params)

ltmp_sub_brm2$fit |> stan_ac(pars = params)

ltmp_sub_brm2$fit |> stan_rhat()

ltmp_sub_brm2$fit |> stan_ess()

ltmp_sub_brm2 |> pp_check(type = 'dens_overlay', ndraws=100) 

ltmp.resids <- make_brms_dharma_res(ltmp_sub_brm2, integerResponse = TRUE)
testUniformity(ltmp.resids)
plotResiduals(ltmp.resids, quantreg = FALSE, asFactor = FALSE)
testDispersion(ltmp.resids)
#nothing super obviously wrong
```
:::

# Model investigation

::: {.panel-tabset}
```{r}
#| label: summary
#| eval: true
#| echo: true

#not really needed because we are just interested in trends
#but we will look and see where the most variation is
ltmp_sub_brm2 |> summary()
#there is more variability at the transect level (0.2) than the site level (0.12)
```

```{r}
#| label: plot
#| eval: true
#| echo: true

#proportion of coral in each year
ltmp_sub_brm2 |> 
  emmeans(~fYEAR, type = "response") |> 
  as.data.frame() |> 
  mutate(Year = as.numeric(as.character(fYEAR))) |> 
  ggplot(aes(y = prob, x = Year)) +
  geom_ribbon(aes(ymin = lower.HPD, ymax = upper.HPD), 
              fill = "orange", 
              alpha = 0.4) +
  geom_line(color = "black") +
  geom_point() +
  scale_y_continuous("Hard coral cover (%)", labels = function(x) x*100) +
  theme_classic()
```

```{r}
#| label: practice in making a function in R
#| eval: true
#| echo: true

SE <- function(x) {
  sd(x) / sqrt(length(x))
}

SE(1:5)
```

```{r}
#| label: comparisons between years
#| eval: true
#| echo: true

newdata <- data.frame(fYEAR = c(2023, 2017))

ltmp_sub_brm2 |> 
  emmeans(~fYEAR, at = newdata, type = "response") |> 
  pairs()

#coral over is 3.77 times higher in 2023 than 2017 --> 277% higher (3.77-1)*100

#when there's a single value --> need to log --> becomes %
#when there are 2 things/a slope/a difference between 2 things --> need to exp --> becomes fractional/odds scale
```
:::


# Further investigations

::: {.panel-tabset}

:::


```{r}
#| label: as a gam
#| results: markup
#| eval: true
#| echo: true
#| cache: false

ltmp.form <- bf(
    n.points | trials(total.points) ~ s(REPORT_YEAR, bs = "tp", k = 10) + #identifying thin plate spline and maximum number of knots
        (1 | SITE_NO) + (1 | TRANSECT_NO),
    family = binomial(link = "logit"))

get_prior(ltmp.form, data = ltmp_sub)
#intercept for continuous variable is the value of y when x = 0 (average coral cover for the average year)
#when you ask for a spline, the data is automatically centered and scaled

ltmp_sub |>
    summarise(
        Median = median(qlogis(HC/100)),
        MAD = mad(qlogis(HC/100)),
        N = mean(total.points)
    )

priors <- prior(normal(-1.1, 0.6), class = "Intercept") +
    prior(normal(0, 1), class = 'b') +
    prior(student_t(3, 0, 0.6), class = 'sd') + #use same MAD as Intercept
    prior(student_t(3, 0, 15), class = 'sds') #MAD is usually half of the x axis (so if 30 years, then do 15)

ltmp_sub.brm3 <- brm(ltmp.form,
                  data = ltmp_sub,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 1000,
                  chains = 3, cores = 3,
                  thin = 10,
                  refresh = 100,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  backend = "cmdstanr",
                  seed = 1
                  )
## save(ltmp_sub.brm4, file = '../ws/testing/ltmp_sub.brm4')
ltmp_sub.brm3 |>
  conditional_effects(conditions = data.frame(total.points = 200)) |>
  plot(points = TRUE)
#GAMs smooth over a lot!! Missing some of our coral declines and recovery
#They are also know to struggle over the ends making them untrustworthy and trends are not stable
#Gaps are also not handled well
```

```{r}
#| label: as a gp
#| results: markup
#| eval: true
#| echo: true
#| cache: false

#gp = gaussian process
#it's great with gaps unlike gam
ltmp.form <- bf(
    n.points | trials(total.points) ~ gp(REPORT_YEAR, k = NA, cov = "exp_quad") +
        (1 | SITE_NO) + (1 | TRANSECT_NO),
    family = binomial(link = "logit"))

#priors
get_prior(ltmp.form, data = ltmp_sub)

invgamma_from_weak_distance(
  lower_distance = 0.10,
  upper_distance = 0.5,
  weak_correlation = 0.10,
  probability = 0.90
)

standist::visualize("inv_gamma(4.6, 0.4", xlim = c(0,2))

priors <- prior(normal(-1.1, 0.6), class = "Intercept") +
    prior(student_t(3, 0, 0.6), class = 'sd') + #use same MAD as Intercept
    prior(inv_gamma(4.6, 0.4), class = 'lscale', coef = "gpREPORT_YEAR") + #how quickly do I except correlation to decline
    prior(student_t(3, 0, 0.6), class = 'sdgp') #same as sd

ltmp_sub.brm4 <- brm(ltmp.form,
                  data = ltmp_sub,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 1000,
                  chains = 3, cores = 3,
                  thin = 10,
                  refresh = 100,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  backend = "cmdstanr",
                  seed = 1
                  )

#raw data
ltmp_sub.brm4 |>
  conditional_effects(conditions = data.frame(total.points = 200)) |>
  plot(points = TRUE)

#proportion of hard coral cover
ltmp_sub.brm4 |>
  conditional_effects(conditions = data.frame(total.points = 1))
```

```{r}
#| label: diagnostics
#| eval: true
#| echo: true

pars <- ltmp_sub.brm4 |> 
  get_variables() |> 
  str_subset("^b_.*|^sd.*")
pars

#ltmp_sub.brm4 |> SUYR_prior_and_posterior()
#now do this
ltmp_sub.brm4$fit |> stan_trace(pars = pars)

ltmp_sub.brm4$fit |> stan_ac(pars = pars)

ltmp_sub.brm4$fit |> stan_rhat()

ltmp_sub.brm4$fit |> stan_ess()

ltmp_sub.brm4 |> pp_check(type = 'dens_overlay', ndraws=100) 

ltmp.resids <- make_brms_dharma_res(ltmp_sub.brm4, integerResponse = TRUE)
testUniformity(ltmp.resids)
plotResiduals(ltmp.resids, quantreg = FALSE, asFactor = FALSE)
testDispersion(ltmp.resids)
#nothing super obviously wrong, could increase iterations
```

```{r}
#| label: predict
#| eval: true
#| echo: true

#we want to predict to 2030
newdata <- data.frame(REPORT_YEAR = seq(1995, 2030, len = 100), SITE_NO = NA, TRANSECT_NO = NA, 
                      total.points = 1)
head(newdata)

ltmp_sub.brm4 |> 
  epred_draws(newdata = newdata, re_formula = NA) |> 
  ungroup() |> 
  dplyr::select(-SITE_NO, -TRANSECT_NO, -.chain, -.iteration, -.row, -total.points) |> 
  group_by(REPORT_YEAR) |> 
  summarise_draws(median, HDInterval::hdi)

ltmp_sub.brm4 |> 
  epred_draws(newdata = newdata, re_formula = NA) |> 
  ungroup() |> 
  dplyr::select(-SITE_NO, -TRANSECT_NO, -.chain, -.iteration, -.row, -total.points) |> 
  group_by(REPORT_YEAR) |> 
  summarise_draws(median, HDInterval::hdi) |> 
  ggplot(aes(y = median, x = REPORT_YEAR)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              fill = "orange", 
              alpha = 0.4) +
  geom_line(color = "black") +
  geom_point() +
  scale_y_continuous("Hard coral cover (%)", labels = function(x) x*100) +
  theme_classic()
```
```{r}
#| label: forecast plot
#| eval: true
#| echo: true

ltmp_sub.brm4 |> 
  epred_draws(newdata = newdata, re_formula = NA) |> 
  ungroup() |> 
  dplyr::select(.draw, REPORT_YEAR, .epred) |> 
  mutate(flag = ifelse(REPORT_YEAR > 2023, "forecast", "historical")) |> 
  ggplot(aes(y = .epred, x = REPORT_YEAR)) +
  geom_line(aes(group = .draw, color = flag), alpha = 0.1) +
  geom_point(data = ltmp_sub,
             aes(y = HC/100, x = REPORT_YEAR))
```

```{r}
#| label: comaprisons
#| eval: true
#| echo: true

newdata <- data.frame(REPORT_YEAR = c(2017, 2023), SITE_NO = NA, TRANSECT_NO = NA, total.points = 1)

ltmp_sub.brm4 |> 
  epred_draws(newdata = newdata, re_formula = NA) |> 
  #filter(.draw == 1) |> 
  ungroup() |> 
  group_by(.draw) |> 
  summarise(diff = diff(.epred)) |> 
  summarise_draws(median, HDInterval::hdi, Pl = ~mean(.x<0), Pg = ~mean(.x>0))

#coral cover increases by 20% units from 2017-2023
#this is on absolute scale/response scale

#to get fold scale (ratio)
#The core trick here is using logarithms (log) and exponentials (exp) to turn a subtraction (difference) into a division (ratio). In mathematics, subtracting logs is the same as dividing the original numbers: \(\log(B) - \log(A) = \log(B / A)\). By taking the exponential of that result, you get \(B / A\), which is the fold-change.
ltmp_sub.brm4 |> 
  epred_draws(newdata = newdata, re_formula = NA) |> 
  ungroup() |> 
  group_by(.draw) |> 
  mutate(.epred = log(.epred)) |> #log scale
  summarise(diff = diff(.epred)) |> #takes log difference --> aka log(B/A)
  mutate(diff = exp(diff)) |> #exp to get B/A ratio (removes log)
  summarise_draws(median, HDInterval::hdi, Pl = ~mean(.x<1), Pg = ~mean(.x>1))

#coral cover has increased 2.77 times
```
