---
title: "Bayesian GLMM Part5"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(knitr)     #for kable
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
library(standist)   #for visualizing distributions
library(cmdstanr)   #for cmdstan
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)     #framework for stats, modelling and visualisation
library(patchwork)
library(geoR)
source('helperFunctions.R')
```

# Scenario

Some ornithologists were interested in the degree of sibling negotiations in owl
chicks.  Specifically, they wanted to explore how sibling negotiations were
affected by feeding satiety and the sex of the parent returning to the nest.
The ornithologists had accessed to a number of owl nests and were able to count
(via recording equipment) the number of sibling negotiations (calls) that the
owl chicks made when the parent returned to the nest.

We could hypothesise that the chicks might call more if they were hungry.  As
part of the investigation, the researchers were able to provided supplementary
food.  As such, they were able to manipulate the conditions such that sometimes
the chicks in a nest would be considered deprived of supplementary food and at
other times they were satiated.

As a parent returned, the researchers recorded the number of sibling
negotiations (calls) along with the sex of the parent.  Since the number of
calls is likely to be a function of the number of chicks (the more chicks the
more calls), the researchers also counted the number of siblings in the brood.

Each nest was measured on multiple occasions.  Hence, we must include the nest
as a random effect to account for the lack of independence between observations
on the same set of siblings.

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls <- read_csv("../data/owls.csv", trim_ws = TRUE)

str(owls)

#need an offset in the model because we have counts (Poisson or neg binomial) but then we are dividing by broodsize which means we won't have integers and cant use poisson or neg binomial. The offset fixes this

#Nest
#FoodTreatment = Cat
#SexParent = Cat
#SiblingNegotiation = counts, numeric
#BroodSize = numeric
#NegPerChick = numeric

#always start with random effects
# NEST (random effect)
# FoodTreatment (fixed effect)
# SexParent (fixed effect)
# FoodTreatment:SexParent (fixed effect)
```


# Data preparation
```{r}
#| label: data prep
#| eval: true
#| echo: true

owls <- owls |> 
  mutate(Nest = factor(Nest),
         FoodTreatment = factor(FoodTreatment),
         SexParent = factor(SexParent),
         NCalls = SiblingNegotiation)
```

# Exploratory data analysis

```{r}
#| label: boxplot
#| eval: true
#| echo: true

owls |> ggplot(aes(y = NegPerChick, x = FoodTreatment, fill = SexParent)) +
  geom_violin() +
  geom_point(aes(color = SexParent), position = position_jitterdodge(jitter.width = 0.3, dodge.width = 1)) +
#except poisson to be skewed so this is okay
#lots of zeros but don't know if thats a bad thing yet
  
#look at nests
owls |> ggplot(aes(y = NegPerChick, x = FoodTreatment, fill = SexParent)) +
  geom_violin() +
  geom_point(aes(color = SexParent), position = position_jitterdodge(jitter.width = 0.3, dodge.width = 1)) +
  facet_wrap(~Nest)

#nests are different so maybe random slopes
```

Random intercept/slope model formula:

$$
\begin{align}
y_i &\sim{} \mathcal{NB}(\mu_{ij}, \phi)\\
ln\left(\mu_{ij}\right) &=\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}\\
\beta_0 &\sim{} \mathcal{N}(0.4, 0.7)\\
\beta_1 &\sim{} \mathcal{N}(0, 1)\\
\boldsymbol{\gamma_j} &\sim{} \mathcal{N}(0, \boldsymbol{\Sigma})\\
\boldsymbol{\Sigma} &= \boldsymbol{D}({\sigma_n^2})\boldsymbol{\Omega}\boldsymbol{D}({\sigma_n^2})\\
\boldsymbol{\Omega} &\sim{} LKJ(\zeta)\\
\sigma_n^2 &\sim{} \mathcal{t}(3,0,0.7)\\
\phi &\sim{} \mathcal{\Gamma}(0.01, 0.01)\\
\end{align}
$$




where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed
and random effects parameters respectively and $\bf{X}$ is the model matrix
representing the overall intercept and effects of food treatment, sex of parent,
arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.  $\bf{Z}$ represents a cell means
model matrix for the random intercepts associated with individual nests.


# Fit the model

This model has the potential to be more complex than most of the
models in previous worksheets. Therefore, we will use the current
example to illustrate the process of building a model up gradually
rather than attempting to fit a more complex model from the start. The
benifit of gradually building a model is that it becomes much easier
to construct more sensible and appropriate models that perform nicely
and meet all the various assumptions.

As we intend to build a number of models, we will nest the MCMC
sampling diagnostics and Model validation sections under the Fit the
model section since each model will have its own associated diagnostic
testing.

Typically, when we are building up models, we start by constructing a
model that only contains the hierarchical (random effects) structure.

```{r}
#| label: fit frequentest model
#| eval: true
#| echo: true

library(glmmTMB)
#start with a simple frequentest model, less time intensive 

#logged broodsize because of the link function we are going to apply
owls_glmm <- glmmTMB(NCalls ~ 1 + offset(log(BroodSize)) + (1 | Nest), data = owls, family = poisson(link = "log"))

#residual diagnostics
resid <- owls_glmm |> simulateResiduals(integerResponse = TRUE)
testUniformity(resid)
plot(resid) #looks like it might be overdispersed
testDispersion(resid) #yeah overdispersed
```

```{r}
#| label: fit bayesian model
#| eval: true
#| echo: true

owls_form <- bf(NCalls ~ offset(log(BroodSize)) + (1 | Nest), family = poisson(link = "log"))

#get priors
get_prior(owls_form, data = owls)
#intercept is the average number of calls per chick
#get average
owls |> summarise(Median = median(log(NegPerChick + 0.5)), MAD = mad(log(NegPerChick + 0.5)))

#set priors
priors <- prior(normal(0.5, 1), class = "Intercept") +
  prior(student_t(3, 0, 1), class = "sd")

owls_brm0 <- brm(owls_form,
                 data=owls,
                 prior=priors,
                 sample_prior="only",
                 iter=3000,
                 chains=3, cores=3, #one core per chain
                 warmup=1000,
                 thin=5,
                 backend="cmdstanr",
                 refresh=0,
                 seed = 1,
                 control = list(adapt_delta = 0.99))

owls_brm0 |> 
  emmeans(~1, type = "response") |> 
  as.data.frame() |> 
  ggplot(aes( x= 1, y = rate)) +
  geom_pointrange(aes(ymin = lower.HPD, ymax = upper.HPD), color = "blue") +
  geom_point(data = owls, aes(y = NegPerChick),
             position = position_jitter(width = 0.2)) +
  scale_y_continuous(trans = scales::pseudo_log_trans())
#priors may be too narrow
#check with data
owls_brm1 <- update(owls_brm0, sample_prior = "yes", seed = 123)
owls_brm1 |> SUYR_prior_and_posterior()
#priors are fine
```
```{r}
#| label: model diagnostics
#| eval: true
#| echo: true

#MCMC diagnosits might not be helpful until the final model, this one isnt built up yet
#residuals are important tho

owls_brm1 |> 
  SUYR_prior_and_posterior()

owls_brm1$fit |> stan_trace(inc_warmup = TRUE)

owls_brm1$fit |> stan_ac()

owls_brm1$fit |> stan_rhat()

owls_brm1$fit |> stan_ess() #ESS is low

owls_brm1 |> pp_check(type = 'dens_overlay', ndraws=100) #not looking good

owls.resids <- make_brms_dharma_res(owls_brm1, integerResponse = TRUE)
testUniformity(owls.resids)
plotResiduals(owls.resids, quantreg = FALSE)
testDispersion(owls.resids)
#big problem with the distribution --> model structure or Poisson 
#check 0 inflation
testZeroInflation(owls_brm1) # we have a zero problem!!

```

### Model 2

**Zero-inflated vs hurdle models**

- zero-inflated models: are a mixture of Bernoulli and Poisson (or negative
  binomial) distributions to model situations where it is believed that the
  observed data are the result of two combined processes (a count process that
  governs how many items there are to count and a Bernoulli process that governs
  whether the items are detectable).  In this way, zero-inflated models are
  useful in situations where we believe that some of the zeros are false zeros
  (recorded as zeros when they should not have been).
- hurdle models: are for situations where the response itself is thought to be
  the result of two processes: one that governs whether there is a response and
  then another that governs the value of the response (must be positive).

:::: {.panel-tabset}

#### Fit model
```{r}
#| label: fit zero inflated model
#| eval: true
#| echo: true

owls_form <- bf(NCalls ~ offset(log(BroodSize)) + (1|Nest), #modeling detectablility
                zi = ~1, #avergae amount of detectablity across whole experiment, could put something else here like turbidity or something that would affect detection
                family = zero_inflated_poisson(link = "log"))

owls_form |> 
  get_prior(data = owls)
#need to include zi Intercept
priors <- prior(normal(0.5, 1), class = "Intercept") +
  prior(student_t(3, 0, 1), class = "sd") +
  prior(normal(0, 1), class = "Intercept", dpar = "zi")
```

```{r}
#| label: run zero inflated model
#| eval: true
#| echo: true

owls_brm2 <- brm(owls_form,
                 data=owls,
                 prior=priors,
                 sample_prior="yes",
                 iter=3000,
                 chains=3, cores=3, #one core per chain
                 warmup=1000,
                 thin=10,
                 backend="cmdstanr",
                 refresh=0,
                 seed = 1,
                 control = list(adapt_delta = 0.99))

```

```{r posterior2h_2c, results='markdown', eval=TRUE, fig.width = 7, fig.height = 5}
owls_brm2 |> SUYR_prior_and_posterior()
```

```{r}
#| label: model diagnostics again
#| eval: true
#| echo: true

#MCMC diagnosits might not be helpful until the final model, this one isnt built up yet
#residuals are important tho

owls_brm2 |> 
  SUYR_prior_and_posterior()

owls_brm2$fit |> stan_trace(inc_warmup = TRUE)

owls_brm2$fit |> stan_ac()

owls_brm2$fit |> stan_rhat()

owls_brm2$fit |> stan_ess() #ESS is low

owls_brm2 |> pp_check(type = 'dens_overlay', ndraws=100) #not looking good

owls.resids <- make_brms_dharma_res(owls_brm2, integerResponse = TRUE)
testUniformity(owls.resids)
plotResiduals(owls.resids, quantreg = FALSE)
testDispersion(owls.resids)
#check 0 inflation
testZeroInflation(owls_brm2) # zeros look fine but other areas are off

```
```{r}
#| label: add in fixed effects for frequentest
#| eval: true
#| echo: true

owls_glmm3 <- glmmTMB(NCalls ~ FoodTreatment * SexParent + offset(log(BroodSize)) + (1 | Nest), 
                      ziformula = ~1, data = owls, family = poisson(link = "log"))

resids <- owls_glmm3 |> simulateResiduals(integerResponse = TRUE)
plot(resids)
#still an issue with the structure of this model (included factors etc)
#could be three things 1) need to include random effects, 2) assumed detectibility was a constant but might be wrong, 3) perhaps overdispersion

#maybe dectectability is skewed, include food treatment and parent
owls_glmm4 <- glmmTMB(NCalls ~ FoodTreatment * SexParent + offset(log(BroodSize)) + (1 | Nest), 
                      ziformula = ~ FoodTreatment * SexParent, 
                      data = owls, family = poisson(link = "log"))
resids <- owls_glmm4 |> simulateResiduals(integerResponse = TRUE)
plot(resids) #looks slightly better
owls_glmm4 |> summary()
#intercept is detectability of female deprived group
plogis(-1.3696) #=20% false 0s
exp(-1.3696) #odds of being false 0s is 1:4

#estimates in zero inflated model have to be exp because they are all of log scale
exp(1.13) #odds are 3x higher when satiated

```
```{r}
#| label: frequentest model w updated structure
#| eval: true
#| echo: true

owls_glmm5 <- glmmTMB(NCalls ~ FoodTreatment*SexParent + offset(log(BroodSize)) +
                        (FoodTreatment*SexParent | Nest),
                      ziformula = ~ FoodTreatment*SexParent, data = owls, family = poisson(link = "log"))
resids <- owls_glmm5 |> simulateResiduals(integerResponse = TRUE)
plot(resids)
testDispersion(resids)
```
```{r}
#| label: bayesian model w updated structure
#| eval: true
#| echo: true

owls_form <- bf(NCalls ~ FoodTreatment*SexParent + 
                  offset(log(BroodSize)) + 
                  (FoodTreatment*SexParent|Nest), 
                zi = ~ FoodTreatment*SexParent,
                family = zero_inflated_poisson(link = "log"))

owls_form |> get_prior(data = owls)

owls |> dplyr::group_by(FoodTreatment, SexParent) |> 
  summarise(Median = median(log(NegPerChick + 0.5)), MAD = mad(log(NegPerChick + 0.5)))
```
```{r}
#| label: priors
#| eval: true
#| echo: true

priors <- prior(normal(0.7, 1), class = "Intercept") +
  prior(normal(0,1), class = "b") +
  prior(student_t(3, 0, 1), class = "sd") +
  prior(lkj_corr_cholesky(1), class = "cor") +
  prior(normal(0, 1), class = "Intercept", dpar = "zi") +
  prior(normal(0, 1), class = "b", dpar = "zi")
```

```{r}
#| label: run hopefully final model
#| eval: true
#| echo: true

owls_brm6 <- brm(owls_form,
                 data=owls,
                 prior=priors,
                 sample_prior="yes",
                 iter=3000,
                 chains=3, cores=3, #one core per chain
                 warmup=1000,
                 thin=10,
                 backend="cmdstanr",
                 refresh=0,
                 seed = 1,
                 control = list(adapt_delta = 0.99))
```
#### MCMC sampling diagnostics
```{r}
#| label: model diagnostics again
#| eval: true
#| echo: true

#MCMC diagnosits might not be helpful until the final model, this one isnt built up yet
#residuals are important tho

owls_brm6$fit |> stan_trace(inc_warmup = TRUE)

owls_brm6$fit |> stan_ac()

owls_brm6$fit |> stan_rhat()

owls_brm6$fit |> stan_ess() #ESS is low

owls_brm6 |> pp_check(type = 'dens_overlay', ndraws=100) #not looking good

owls.resids <- make_brms_dharma_res(owls_brm6, integerResponse = TRUE)
testUniformity(owls.resids)
plotResiduals(owls.resids, quantreg = FALSE)
testDispersion(owls.resids)
#check 0 inflation
testZeroInflation(owls_brm6) # zeros look fine but other areas are off

owls_brm6 |> 
  conditional_effects("FoodTreatment:SexParent") |> 
  plot(points = TRUE)

get_variables(owls_brm6)
```
```{r}
#| label: summary table
#| eval: true
#| echo: true

owls_brm6 |> 
  as_draws_df() |> 
  dplyr::select(matches("^b_.*|^sd_.*")) |> 
  mutate(across(matches("^b_.*"), exp)) |> 
  summarise_draws(median, 
                  HDInterval::hdi,
                  rhat, 
                  ess_bulk,
                  Pl = ~mean(.x < 1), #less than 
                  Pg = ~mean(.x > 1)) #greater than

#b_Intercept = average of 2.3 calls per chick in the deprived female group, BUT conditional effects plot shows calls per NEST
#b_FoodTreatmentSatiated = difference between female deprived and female satiated. Calls per chick in satiated group are halved
# no evidence of interaction effects
```



### Comparisons
```{r}
#| label: comparisons
#| eval: true
#| echo: true

#this gives us average calls per chick
owls_brm6 |> emmeans(~FoodTreatment*SexParent, type = "response")

# the | runs FoodTreatment separately for each SexParent
owls_brm6 |> emmeans(~FoodTreatment | SexParent, type = "response")

#pairs() compares things, so its comparing all combinations
owls_brm6 |> 
  emmeans(~FoodTreatment*SexParent, type = "response") |> 
  pairs()

#but when you do it this way you get FoodTreatment compared separately for each SexParent
#think of SexParent here as the grouping by variable
#these are on a fractional scale because type = "response" back transforms 
owls_brm6 |> 
  emmeans(~FoodTreatment | SexParent, type = "response") |> 
  pairs()

#if we add regrid before the pairs, then the transformation occurs before subtraction and we can get absolute values/differences
owls_brm6 |> 
  emmeans(~FoodTreatment | SexParent, type = "response") |> 
  regrid() |> 
  pairs()
```

Conclusions:
- evidently, the zero-inflated negative binomial model with random
  intercepts/slopes (owls_brm5b) is the best model...

:::
<!-- END_PRIVATE-->


# Partial effects plots


# Model investigation

# Further investigations


# References
