---
title: "Regression trees example 3"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, tidy='styler')
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(gbm)         #for gradient boosted models
## library(gbm3)         #for gradient boosted models
#install.packages("pdp")
library(car)
library(pdp)
library(ggfortify)
#install.packages("randomForest")
library(randomForest)
library(tidyverse)
library(patchwork)
```


# Scenario
@newman-2015 sought to investigate the relative importance of reef habitat
complexity and structural components of the benthos to the total
richness of fish species in Caribbean reefs. To do so, they recorded
they surveyed fish and benthos from one hundred 25m^2 plots within
three marine reserves. Plots were haphazardly selected within three
Countries (St Vincent and the Grenadines, Puerto Rico and Bonaire) and
to represent a range of structural and complexity conditions which
comprised:

- the number of corals
- the number of tall corals
- slope angle
- maximum sponge height
- maximum octocoral height
- reef complexity


Format of newman.csv data file

| total_rich | Reef_complexity | No_tall_corals | No_corals | Sponge_max_height | Octocoral_max_height | Slope_angle | Country |
|------------|-----------------|----------------|-----------|-------------------|----------------------|-------------|---------|
|            |                 |                |           |                   |                      |             |         |
| 25         | 1               | 0              | 14        | 5                 | 13                   | 0           | SVG     |
| 39         | 1               | 0              | 17        | 6                 | 24                   | 0           | SVG     |
| 52         | 1               | 0              | 28        | 12                | 15                   | 0           | SVG     |
| 42         | 1               | 0              | 17        | 11                | 121                  | 5           | SVG     |
| 51         | 1               | 0              | 23        | 13                | 27                   | 5           | SVG     |
| 51         | 1               | 0              | 21        | 38                | 18                   | 0           | SVG     |
| 49         | 1               | 0              | 22        | 31                | 91                   | 5           | SVG     |
| 50         | 1               | 0              | 12        | 14                | 47                   | 5           | SVG     |
| 51         | 1               | 1              | 20        | 34                | 29                   | 0           | SVG     |
| 52         | 2               | 2              | 114       | 0                 | 0                    | 25          | SVG     |



# Read in the data

```{r readData, results='markdown', eval=TRUE}
newman <- read_csv('../data/newman.csv', trim_ws=TRUE)
glimpse(newman)
```
# Prepare data
```{r}
#| label: prep data
#| eval: true
#| echo: true

#remove spaces in variable names
#take all names and replace spaces with _
names(newman) <- str_replace_all(names(newman), " ", "_")
names(newman)

#declare categorical variables
newman <- newman |> 
  mutate(
    fCountry = factor(Country),
    fReef_complexity = factor(Reef_complexity)
  )
```

# Exploratory data analysis
```{r}
#| label: scatterplot
#| eval: true
#| echo: true

scatterplotMatrix(~total_rich + Reef_complexity + No_tall_corals +
                    No_corals + Sponge_max_height + Octocoral_max_height +
                    Slope_angle + fCountry, data = newman,
                  diagonal = list(method = 'boxplot'))

#dont use assumptions like we normally do
#dont need to transform variables

#but do need to decide what loss function to use
# we have count data --> Poisson distribution --> log --> Poisson loss function

#if you want to know what variables are most important and get a partial effects plot, then you shouldn't put correlated predictors into the same regression tree

#enforce monotonic trends on anything?
# look at top row
#never impose monotonic trend on categorical variables
#the next 3 seem positive, octocoral height leave 0, slope positive, and remove country because it will likely outcompete the other variables

#every predictor is either positive (1), none (0), or negative (-1)
```

# Fit the model
```{r}
#| label: fit model
#| eval: true
#| echo: true

#all predictors are additive, no interactions because gbm automatically will include interactions
newman_gbm <- gbm(fish_rich ~ fReef_complexity +
                    No_tall_corals +
                    No_corals +
                    Sponge_max_height +
                    Octocoral_max_height +
                    Slope_angle,
                  data = newman,
                  distribution = "poisson", #loss function
                  n.trees = 10000, #a good starting point
                  var.monotone = c(0, 1, 1, 1, 0, 1), #monotonic trends
                  interaction.depth = 5, #how many splits does each tree have (typically 5, 7, or 9)
                  shrinkage = 0.001, #learning rate, only going to retain 1% in memory
                  bag.fraction = 0.5, #50% of data are in and 50% are out for testing at tree level
                  train.fraction = 1, #determines how much is completely withheld and never seen by the model for training (1 = none withheld)
                  n.minobsinnode = 10, #node cannot have less than 10 data points
                  cv.folds = 10 #how many times should each tree be repeated
                  )

#need to determine when to cutoff models (aka optimal number of trees)

#Option 1
(best.iter <- gbm::gbm.perf(newman_gbm, method = 'OOB'))
#there is a single number print out that tells you the optimal trees
#in this case 102 trees
#but we think it learned too fast, so now we have to tell it to redo it
#NEED to achieve 1000 minimum trees

#Option 2
(best.iter <- gbm::gbm.perf(newman_gbm, method = 'cv'))
#says only 30 trees, so even worse
#green line represents how well it is predicting, want the dip above 1000

#need to make model run slower! Changing shrinkage from 0.1 to 0.001

#now we get 1965 trees and 2981 trees!
```

# Explore relative influence
```{r}
#| label: relative influence
#| eval: true
#| echo: true

summary(newman_gbm, n.trees = best.iter)
#reef complexity is the most important

#we have 6 variables so if we split up the 100% by 6 we see what variables have a higher % and they are therefore better than random
100/6
#anything above 16.67 is important/justifiable to keep
```

# Explore partial effects

```{r}
#| label: partial effects plots
#| eval: true
#| echo: true


attr(newman_gbm$Terms, "term.labels")

plot(newman_gbm, 1, n.tree = best.iter)
#as reef complexity increases, fish richness increases

plot(newman_gbm, 5, n.tree = best.iter)
#as octocoral height increases, fish richness increases and then decreases

#a nicer way to plot these
newman_gbm |> 
  pdp::partial(pred.var = 'Octocoral_max_height',
               n.trees = best.iter,
               type = "regression",
               recursive = FALSE,
               ice = TRUE,
               inv.link = exp) |> 
  autoplot()

#but all of these plots are missing uncertainty, have to add ice = TRUE to get the machine learning community's version of uncertainty...but its not true uncertainty

#other option is bootstrapping
```

# Bootstrapping
```{r}
#| label: bootstrapping
#| eval: true
#| echo: true

#goal is to create a partial plot for Octocoral_max_height

#Create 2 lists:
 # - relative_influence_list
 # - prediction_list

#Need to create newdata (dataset for predictions)

#Step 1: Resample the data
#Step 2. Fit a tree
#Step 3. Cross validation (figure out optimal number of trees)
#Step 4. Calculate relative influence (how important was each predictor)
#Step 5. Predict

#Compile relative influence -> plot
#Compile predictions -> plot

#and we want to do this multiple times (10x)
#need to create a loop

nBoot <- 10 #number of bootstraps
rel_inf_list <- vector("list", nBoot) #empty list with 10 spaces
pred_list <- vector("list", nBoot)
newdata <- with(newman, data.frame(
  Octocoral_max_height = seq(min(Octocoral_max_height), 
                             max(Octocoral_max_height), len = 100), 
  No_tall_corals = mean(No_tall_corals),
  No_corals = mean(No_corals),
  Sponge_max_height = mean(Sponge_max_height),
  Slope_angle = mean(Slope_angle))) |> 
  crossing(fReef_complexity = levels(newman$fReef_complexity))
                      

for (i in 1:nBoot) {
  print(paste("Bootstrap no.", i))
  
  dat <- newman |> 
    sample_n(size = n(), replace = TRUE)
  
  mod_gbm <- gbm(fish_rich ~ fReef_complexity +
                    No_tall_corals +
                    No_corals +
                    Sponge_max_height +
                    Octocoral_max_height +
                    Slope_angle,
                  data = dat,
                  distribution = "poisson", 
                  n.trees = 10000, 
                  var.monotone = c(0, 1, 1, 1, 0, 1), 
                  interaction.depth = 5, 
                  shrinkage = 0.001, 
                  bag.fraction = 0.5, 
                  train.fraction = 1, 
                  n.minobsinnode = 10, 
                  cv.folds = 10
                  )
  
  best.iter <- gbm::gbm.perf(mod_gbm, method = 'cv')
  
  rel_inf_list[[i]] <- summary(mod_gbm, Boot = i, n.trees = best.iter)
  
  fit <- predict(mod_gbm, newdata = newdata, n.trees = best.iter)
  pred_list[[i]] <- data.frame(newdata, Fit = fit, Boot = i)
  
}

#check
str(rel_inf_list)

#compile relative influences and make a plot
rel_inf <- do.call("rbind", rel_inf_list) #do this, on this (aka take list and bind all the shelves into one)
head(rel_inf)

#rel_inf |> 
#  group_by(var) |> 
#  posterior::summarise_draws(median, HDInterval::hdi)

rel_inf |> 
  group_by(var) |> 
  summarise(median = median(rel.inf), lower = HDInterval::hdi(rel.inf)[1], upper = HDInterval::hdi(rel.inf)[2])

#plot of medians and relative importance
g1 <- rel_inf |> 
  group_by(var) |> 
  summarise(median = median(rel.inf), 
            lower = HDInterval::hdi(rel.inf)[1], 
            upper = HDInterval::hdi(rel.inf)[2]) |> 
  ggplot(aes(x = median, y = reorder(var, median))) + #orders variables by median
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  geom_vline(xintercept = 16.66, linetype = "dashed")
g1

#plot with uncertainty
pred <- do.call("rbind", pred_list)
#Fit is on a log scale though so we need to back transform
pred <- pred |> 
  mutate(Fit = exp(Fit)) |> 
  group_by(fReef_complexity, Octocoral_max_height) |> 
  tidybayes::median_hdci(Fit) #or use summarise_draws
pred

g2 <- pred |> 
  ggplot(aes(y = Fit, x = Octocoral_max_height, fill = fReef_complexity)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_line(aes(color = fReef_complexity))
g2

g2 | g1 #two plots side by side

g1 + patchwork::inset_element(g2, left = 0.4, bottom = 0.01, right = 1, top = 0.4) & theme_classic()

g2 + patchwork::inset_element(g1, left = 0.01, bottom = 0.01, right = 0.5, top = 0.35) & theme_classic()

```

# Try random forest model
```{r}
#| label: random forest code
#| eval: true
#| echo: true

newman.rf <- randomForest(
  fish_rich ~ 
    fReef_complexity +
    No_tall_corals +
    No_corals +
    Sponge_max_height +
    Octocoral_max_height +
    Slope_angle,
  data = newman, 
  importance = TRUE,
  ntree = 1000
)

#extract importance values
newman.imp <- randomForest::importance(newman.rf)
100 * newman.imp / sum(newman.imp)
#%IncMSE = if you left out Octocoral_max_height, you would have the most residuals or unexplained variance, therefore it is the most important
#IncNodePurity = leaving the no_tall_corals out, leads to the most inaccuracies
```
